---
title: "Hotel Booking Cancellation & Demand Analysis"
author: "ANAND M"
date: "`r Sys.Date()`"
output: html_document
---

## 🔹 Business Problem

Hotel businesses experience revenue loss due to booking cancellations and unpredictable demand.  
Understanding the drivers of cancellations and demand trends is crucial for effective revenue management.

## 🔹 Objective

The objective of this case study is to analyze hotel booking data to:

- Identify factors influencing booking cancellations  
- Understand seasonal and hotel-type demand trends  
- Provide actionable recommendations to reduce cancellations and improve revenue

## 🔹 Key Questions

- Which hotel type has higher cancellation rates?  
- How does lead time affect cancellations?  
- Does deposit type reduce cancellation probability?  
- Which months show peak demand?  
- How does ADR influence cancellations?

---

##   Tools & Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(ggplot2)

if(!dir.exists("plots")){
dir.create("plots")
}


hotel_df <- read_csv("C:/Users/anand/Documents/Hotel_Booking_Case_Study/R_PROGRAM/hotel_bookings.csv")
hotel_df <- hotel_df %>% mutate(total_guests = adults + children + babies)
```

##   Cancellation Rate by Hotel Type

```{r cancellation, message=FALSE, warning=FALSE}
hotel_df %>%
group_by(hotel) %>%
summarise(
total_bookings = n(),
cancellations = sum(as.numeric(as.character(is_canceled))),
cancellation_rate = mean(as.numeric(as.character(is_canceled)))*100
)
p1 <- ggplot(hotel_df, aes(x=hotel, fill=is_canceled)) +
geom_bar(position="fill") +
ylab("Proportion") +
ggtitle("Cancellation Rate by Hotel Type") +
scale_fill_manual(values=c("0"="green","1"="red"), labels=c("Not Canceled","Canceled"))
p1
ggsave("plots/cancellation_by_hotel.png", plot=p1, width=6, height=4)
```

Insight:
City hotels show higher cancellation rates compared to resort hotels

##  Lead Time vs Cancellation
```{r leadtime, message=FALSE, warning=FALSE}
hotel_df %>%
group_by(is_canceled) %>%
summarise(avg_lead_time = mean(lead_time))


p2 <- ggplot(hotel_df, aes(x=is_canceled, y=lead_time, fill=is_canceled)) +
geom_boxplot() +
ylab("Lead Time (Days)") +
xlab("Canceled") +
ggtitle("Lead Time vs Cancellation") +
scale_fill_manual(values=c("0"="green","1"="red"))
p2
ggsave("plots/leadtime_vs_cancellation.png", plot=p2, width=6, height=4)
```

Insight:
Longer lead times are associated with higher likelihood of cancellations.

##  Monthly Booking Demand

```{r monthly, message=FALSE, warning=FALSE}
monthly_demand <- hotel_df %>%
group_by(arrival_date_month) %>%
summarise(total_bookings = n())
monthly_demand
p3 <- ggplot(monthly_demand, aes(x=arrival_date_month, y=total_bookings, group=1)) +
geom_line(color="blue", size=1.2) +
geom_point(color="red", size=2) +
xlab("Month") + ylab("Total Bookings") +
ggtitle("Monthly Booking Demand Trend") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
p3
ggsave("plots/monthly_booking_demand.png", plot=p3, width=6, height=4)
```
Insight:
Peak bookings occur in July and August; January and February show low demand.


##  Deposit Type vs Cancellation

```{r deposit, message=FALSE, warning=FALSE}
hotel_df %>%
group_by(deposit_type) %>%
summarise(cancellation_rate = mean(as.numeric(as.character(is_canceled)))*100)
p4 <- ggplot(hotel_df, aes(x=deposit_type, fill=is_canceled)) +
geom_bar(position="fill") +
ylab("Proportion") +
ggtitle("Cancellation Rate by Deposit Type") +
scale_fill_manual(values=c("0"="green","1"="red"), labels=c("Not Canceled","Canceled"))
p4
ggsave("plots/deposit_type_vs_cancellation.png", plot=p4, width=6, height=4)
```
Insight:
Non-refundable deposits significantly reduce cancellations


##  ADR vs Cancellation
```{r ADR, message=FALSE, warning=FALSE}
p5 <- ggplot(hotel_df, aes(x=is_canceled, y=adr, fill=is_canceled)) +
geom_boxplot() +
ylab("Average Daily Rate (ADR)") +
xlab("Canceled") +
ggtitle("ADR vs Cancellation") +
scale_fill_manual(values=c("0"="green","1"="red"))
p5
ggsave("plots/adr_vs_cancellation.png", plot=p5, width=6, height=4)
```
Insight:
Extreme ADR values can increase the risk of cancellation

##  Total Guests vs Cancellation

```{r totalguest, message=FALSE, warning=FALSE}
p6 <- ggplot(hotel_df, aes(x=is_canceled, y=total_guests, fill=is_canceled)) +
geom_boxplot() +
ylab("Total Guests") +
xlab("Canceled") +
ggtitle("Total Guests vs Cancellation") +
scale_fill_manual(values=c("0"="green","1"="red"))
p6
ggsave("plots/total_guests_vs_cancellation.png", plot=p6, width=6, height=4)
```

Insight:
Total number of guests has some effect on cancellation trends but is less significant than lead time or deposit type.

##  Key Insights

City hotels have higher cancellation rates than resort hotels

Longer lead times increase likelihood of cancellations

Non-refundable deposits reduce cancellations significantly

Peak bookings in July–August, low in January–February

ADR extremes affect cancellations

##  Business Recommendations

Apply stricter cancellation policies for long lead-time bookings

Encourage non-refundable deposit options during peak seasons

Optimize pricing based on seasonal demand

City hotels can focus on corporate/repeat customers to reduce cancellations

##  Conclusion

This case study analyzed 100,000+ hotel bookings to identify cancellation drivers and demand patterns.
Plots have been automatically saved in the plots/ folder for portfolio or GitHub use.
The insights and recommendations can help hotels increase revenue and reduce losses due to cancellations